---
title: "Repeated Measures ANOVA"
author:
  - name: Hannah Bowley
  - name: Micaela Lembo
format: html
toc: true
toc-expand: 6
theme: lux
---

## Repeated measures ANOVA

Repeated measure ANOVA is used for analyzing data where the same subjects are measured more than once

There are three types of repeated measures ANOVA:

1.  **One-way repeated measure ANOVA**: similar to a paired samples t-test.Compares the means of 3 or more categories of 1 within-subject variable (1 independent factor)

2.  **Two-way repeated measure ANOVA**: evaluates the effects of two within-subject factors (2 independent variables) on a continuous dependent variables


### Assumptions

1.  No significant outliers

2.  Normality

3.  Assumption of sphericity

### Hypothesis

(NEED TO ADD IN EQUATION)

**Null hypothesis:** - No difference in group means at any time points

**Alternative hypothesis:** - Significant difference in group mean in at least one or more time points

## Packages and Libraries

### Installing Packages

```{r}
#| echo: true
#| message: false
#| warning: false


# install.packages("tidyverse")
# install.packages("datarium")

```

### Loading Libraries

```{r}
#| echo: true
#| message: false
#| warning: false

library(datarium)
library(tidyverse)
library(rstatix)
library(gt)

```
## One-way Repeated Measures ANOVA

### Looking at data

```{r}
#| echo: true
#| message: false
#| warning: false


data("selfesteem", package = "datarium")

selfesteem

```

The dataset contains 10 individuals' self-esteem score on three time points during a specific diet to determine whether their self-esteem improved.

### Wide to long format

Currently the selfesteem data set is in wide format. In order to run some of the statistical test, we need to transform it into long format

```{r}
#| echo: true
#| message: false
#| warning: false


# 
selfesteem <- selfesteem %>%
  gather(key = "time", value = "score", t1, t2, t3) 

# checking to see if it converted correctly
selfesteem


```


### Run summary statistics

```{r}
#| echo: true
#| message: false
#| warning: false

selfesteem %>% 
  group_by(time) %>% 
  summarize(mean = mean(score),
            sd = sd(score))


```

### Visualizing the data

```{r}
#| echo: true
#| message: false
#| warning: false

bxp <- ggplot(selfesteem, aes(x = time, y = score))+
  geom_boxplot() + 
  ggtitle("Self esteem score across time points") +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))

bxp
```

### Checking Assumptions

#### Normality

```{r}
#| echo: true
#| message: false
#| warning: false

ggplot(selfesteem, aes(sample = score)) +
  stat_qq()+
  geom_qq_line()+
  facet_wrap(~time, ncol = 3)

ggplot(selfesteem, aes(sample = score))+
  stat_qq() +
  geom_qq_line()

```
#### Sphericity
Sphericity: variances of the differences between all combinations of related groups are equal

##### Mauchly's Test of Sphericity 
Null hypothesis: the variances of the differences are equal
Alternative Hypothesis: The variances of the differences are not equal 


### Conducting One-way repeated measures ANOVA

```{r}
#| echo: true
#| message: false
#| warning: false

rm_aov <- anova_test(
  data = selfesteem,
  dv = score,
  wid = id,
  within = time,
  detailed = TRUE
)

rm_aov

gt(get_anova_table(rm_aov))


```
Greenhouse-Geisser sphericity correction applied to within subject factors where the assumptions were violated

### Post-hoc test

```{r}
#| echo: true
#| message: false
#| warning: false

selfesteem %>% 
  pairwise_t_test(score ~  time, paired = TRUE, p.adjust.method =  "bonf") %>% 
  gt()

```
## Two-way Repeated Measures ANOVA

### Looking at data

```{r}
#| echo: true
#| message: false
#| warning: false

head(selfesteem2, 5)
tail(selfesteem2, 5)
```
The selfesteem2 dataset contains the self-esteem score of 12 subjects who were randomized into either a control or special diet treatment. Self-esteem scores were recorded at 3 time points 

### Wide to long format
```{r}
#| echo: true
#| message: false
#| warning: false

selfesteem2_long <- selfesteem2 %>% 
  pivot_longer(cols = c(t1:t3), names_to = "time", values_to = "score")
```

### Run summary statistics

```{r}
#| echo: true
#| message: false
#| warning: false

selfesteem2_long %>% 
  group_by(time, treatment) %>% 
  summarize(mean = mean(score),
            sd = sd(score))

```

### Visualizing the data

```{r}
#| echo: true
#| message: false
#| warning: false

ggplot(
  data = selfesteem2_long,
  aes(x = time, y = score, color = treatment)
  ) + 
  geom_boxplot()
```
### Checking Assumptions

#### Checking for normality visually using qqplot

```{r}
#| echo: true
#| message: false
#| warning: false

ggplot(selfesteem2_long, aes(sample = score)) +
  stat_qq()+
  geom_qq_line()+
  facet_grid(rows = vars(treatment), cols = vars(time))
```
### Conducting Two-way repeated measures ANOVA

```{r}
#| echo: true
#| message: false
#| warning: false

two_way_anova <- selfesteem2_long %>% 
  anova_test(
    dv = score,
    wid = id,
    within = c(treatment, time)
)
two_way_anova

gt(get_anova_table(two_way_anova))

```
### Post-hoc test

```{r}
#| echo: true
#| message: false
#| warning: false

selfesteem2_long %>% 
  group_by(time) %>% 
  pairwise_t_test(score ~ treatment, paired = TRUE, p.adjust.method = "bonf") %>% 
  gt()

```



